---
title: "Final Project Report"
author: "Steven Lautenbach & Mathias Thyssen"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

# 1. Introduction

## Main Question

This project examines whether being a Shooter video game causally increases global sales. Specifically, the question is:

**What is the effect of belonging to the Shooter genre on global sales of video games?**

The treatment variable (X) is an indicator equal to 1 if a game belongs to the Shooter genre, and 0 otherwise. The outcome variable (Y) is the global sales amount, measured in millions of copies sold.

This is a causal and economically meaningful question because we are not simply asking whether Shooter games sell more, but whether being a Shooter game causes higher sales.

## Why this question matters 

This question is relevant to game developers, Publishers, investors, and platform companies such as Sony, Microsoft, and Nintendo. Knowing whether or not a certain genre (in this case shooter games) tends to generate higher sales is truly relevant to the gaming industry because it affects how development resources are allocated, how portfolios are structured, how marketing budgets are set, and even how platform exclusives are planned.

Moreover, understanding whether the genre itself increases sales (holding other factors constant) also helps business decision-makers make more informed strategic choices. If being a Shooter game truly causes higher sales, firms can adjust their investment and planning accordingly.

## Why this question is not obvious

This question is not obvious because simply observing that a certain genre sells more does not mean the genre itself causes higher sales. Several other factors could influence this relationship and are also correlated with whether a game is a Shooter.

Platform differences: Where some can have larger audiences or even favor/support specific genres of games (Which will reflect on the sales for that video game).

Time trends: The gaming market has expanded significantly over time (From 1980s to 2010s). This could also have the idea that there was greater digital distribution of games (more available), and that consumer preferences for certain genres could have changed over time.

Popular franchises: The popularity of certain game titles could also be another variable that impacts sales rather than simply just genre. This is because big game franchises will have loyal fan bases, receive large marketing budgets, and get a lot of repeat sales. More or less, higher sales could reflect more brand recognition rather than genre.

These factors all correlate with the Shooter variable, meaning they can create the appearance that Shooter games sell more even if the genre itself is not the true cause.

## Why descriptive statistics alone are not enough to answer this question and regression  bias explanation.

These factors create the possibility of omitted variable bias: Shooter games may appear to sell more not because of their genre, but because they are disproportionately released on high‑selling platforms, in high‑demand years, or by large publishers. For these reasons, descriptive statistics alone cannot answer the causal question. The goal of this analysis is to explore the data, document the patterns we observe, and identify the confounding factors that must be addressed later in order to credibly estimate the causal effect of being a Shooter game on global sales.


# 2. About the Data

## Data Source

The primary dataset used in this analysis is the Video Game Sales dataset (1980–2016), originally compiled from retail sales data reported by VGChartz, a widely cited video game market tracking firm that aggregates global physical retail sales across major regions (North America, Europe, Japan, and Other regions). The dataset is publicly distributed for academic use via repositories such as: Kaggle and Public GitHub mirrors of the original VGChartz compilation.

The dataset is widely used in academic and applied data analysis projects studying video game market performance.

The dataset covers commercially released console video games between 1980 and 2016.

One notable feature of the dataset is the presence of extreme outliers, most prominently Duck Hunt (1984), which recorded unusually high global sales relative to other titles. This game was bundled with the Nintendo Entertainment System (NES) in many regions, causing its reported “sales” to reflect hardware‑bundle distribution rather than standalone consumer demand. Because of this, Duck Hunt behaves more like a platform‑level anomaly than a typical shooter title. The game is retained in the dataset for transparency, but its influence is evaluated later to ensure it does not distort estimates of the shooter genre’s effect on global sales.

## Sample Coverage

The dataset represents the physical console video game market over this period for video games selling at least 100,000 copies.

## Unit of observation

Each row corresponds to a single game released on a specific platform.

## Number of obsrvations

The dataset contains 16,599 game-platform observations.

## Frequency

Since it is based on release year of each video game, it is yearly data frequency.

## Variables

The dataset contains the following key variables:

Name: The title of the video game. Each observation corresponds to a specific game released on a specific platform.

Genre: The primary genre classification of the game (Action, Shooter, Sports, Role-Playing, etc.). This categorical variable identifies the type of gameplay and is used to construct the treatment indicator for Shooter games.

Publisher: The company responsible for publishing the game (Nintendo, Electronic Arts, Activision). This variable allows us to control for differences in marketing capacity, brand reputation, and distribution networks.

Platform: The gaming console on which the game was released (PlayStation 2, Xbox, Wii). Because platform popularity and install base vary substantially, this variable is an important control for market exposure.

Year: The year in which the game was released. This variable captures time trends in the gaming industry, including overall market growth and changes in consumer preferences.

Sales: Regional and global sales measured in millions of copies sold. The dataset includes: NA_Sales (North America), EU_Sales (Europe), JP_Sales (Japan), Other_Sales, and Global_Sales, which is the sum of regional sales

In this situation, Global_Sales is used as the primary outcome variable.


# 3. Analysis

In this section, we explore the relationship between Shooter games and global sales using descriptive statistics, visualizations, and preliminary regression analysis. The goal is not to estimate a causal effect yet, but to understand what the data suggest and identify the main sources of confounding that must be addressed later.

The analysis proceeds in three steps:

(i) we document basic patterns in global sales for Shooter and non‑Shooter games over time,

(ii) we examine how Shooter games are distributed across platforms, publishers, and years, and how these patterns may bias simple comparisons, and

(iii) we estimate a sequence of regressions that progressively control for platform, year, and publisher differences using a fixed‑effects approach.

Because Shooter games are not randomly assigned across platforms or publishers, the descriptive patterns alone cannot be interpreted causally. The purpose of this section is to show how these underlying differences shape the raw Shooter–sales relationship and to motivate the need for a more careful empirical strategy in the future.



```{r, echo=F}
library(tidyverse)
vgsales <- read_csv("vgsales.csv")
vgsales <- vgsales |>
  mutate(Shooter = ifelse(Genre == "Shooter", 1, 0),
         Shooter = factor(Shooter))
vgsales <- as.data.frame(vgsales)
head(vgsales)
```

## Summary Statistics and Data Overview

```{r, echo=F}
library(stargazer)

summary(vgsales)
stargazer(vgsales, type = "text")

length(unique(vgsales$Publisher))
```

The summary statistics show information about each column in the dataset, showing in particular the different ranges and averages for each region. This also shows that the dataset has 15,288 non-shooter games compared to only 1,310 shooters.

## Plots

These plots summarize mean sales for Shooter and non‑Shooter games, allowing us to compare average performance across years and regions.

### Time-Series Plot

```{r, echo=F}
vgsales |>
  group_by(Year, Shooter) |>
  summarize(Global_Sales = mean(Global_Sales)) |>
  ggplot(aes(as.numeric(Year), Global_Sales,
             color = factor(Shooter, labels = c("No", "Yes")),
             group = Shooter)) +
  geom_line() +
  labs(x = "Year",
       y = "Mean Global Sales (in Millions)",
       color = "Shooter Game?")
```

The graph shows a high degree of sporadic behavior in the first decade recorded by the data, but by the 1990s a general trend emerged with shooter and non-shooter game sales mostly at parity with one another. This trend began to fade in the second half of the 2000s, with shooters steadily increasing their sales lead over non-shooter games. This indicates that time may be a relevant variable to include.

### Regional Sales Plots

```{r, echo=F}
vgsales_long <- vgsales |>
  pivot_longer(cols = NA_Sales:Other_Sales,
               names_to = "Region",
               values_to = "Sales")

vgsales_reg <- vgsales_long |>
  group_by(Year, Region, Shooter) |>
  summarize(Sales = mean(Sales, na.rm = TRUE))

ggplot(vgsales_reg, aes(as.numeric(Year), Sales,
                        color = factor(Shooter, labels = c("No", "Yes")),
                        group = Shooter)) +
  geom_line() +
  facet_wrap(~Region) +
  labs(x = "Year",
       y = "Mean Global Sales (in Millions)",
       color = "Shooter Game?")

# Outlier
vgsales |>
  filter(Name == "Duck Hunt")
```

These graphs show little variation in shooter game sales and relative popularity within or between graphs, indicating that region may not have a significant effect on sales.

## Regression

### Simple Linear Regression

```{r, echo=F}
vg_mod <- lm(Global_Sales ~ Shooter, vgsales)
summary(vg_mod)
```

In a simple linear regression containing only the Shooter dummy variable, the coefficient is found to be statistically significant, providing evidence that a game's status as a shooter or non-shooter game affects sales.

### Multiple Linear Regression

```{r, echo=F}
library(lfe)
vg_mod_pub <- felm(Global_Sales ~ Shooter | Publisher, vgsales)
summary(vg_mod_pub)

vg_mod_plat <- felm(Global_Sales ~ Shooter | Platform, vgsales)
summary(vg_mod_plat)

vg_mod_time <- felm(Global_Sales ~ Shooter | Year, vgsales)
summary(vg_mod_time)

vg_mod_reg <- felm(Global_Sales ~ Shooter | Region, vgsales_long)
summary(vg_mod_reg)

vg_mod_full <- felm(Global_Sales ~ Shooter | Publisher + Platform + Year + Region, vgsales_long)
summary(vg_mod_full)
```

## Log Transformation with Duck Hunt Exclusion

In an effort to remedy some of the extreme values in the data, Duck Hunt was removed and sales were log-transformed in order to elicit a potentially more intuitive visual representation of the trends in the data.

### Visualizations

```{r, echo=F}
vgsales_ndh <- vgsales |>
  filter(Name != "Duck Hunt")

vgsales_ndh |>
  group_by(Year, Shooter) |>
  summarize(Global_Sales = mean(Global_Sales)) |>
  ggplot(aes(as.numeric(Year), log(Global_Sales),
             color = factor(Shooter, labels = c("No", "Yes")),
             group = Shooter)) +
  geom_line() +
  labs(x = "Year",
       y = "Mean Global Sales (in Millions)",
       color = "Shooter Game?")

vgsales_long_ndh <- vgsales_ndh |>
  pivot_longer(cols = NA_Sales:Other_Sales,
               names_to = "Region",
               values_to = "Sales")

vgsales_reg_ndh <- vgsales_long_ndh |>
  group_by(Year, Region, Shooter) |>
  summarize(Sales = mean(Sales, na.rm = TRUE))

ggplot(vgsales_reg_ndh, aes(as.numeric(Year), log(Sales),
                        color = factor(Shooter, labels = c("No", "Yes")),
                        group = Shooter)) +
  geom_line() +
  facet_wrap(~Region) +
  labs(x = "Year",
       y = "Mean Global Sales (in Millions)",
       color = "Shooter Game?")
```

The log-corrected global sales graph more parsimoniously displays a similar version of the trend seen in the original, with sales of both categories being roughly the same prior to 2005, at which time shooter games began superseding non-shooters. The regional graphs also mirror their counterparts but in a much more visually apparent way, with European, North American, and Other sales resembling global patterns while Japan consistently prefers non-shooter games over shooters. This suggests that log-corrections may not drastically change the results of the regression models.

### Regression Models

```{r}
vg_mod_ndh <- lm(log(Global_Sales) ~ Shooter, vgsales_ndh)
summary(vg_mod_ndh)

vg_mod_pub_ndh <- felm(log(Global_Sales) ~ Shooter |
                         Publisher, vgsales_ndh)
summary(vg_mod_pub_ndh)

vg_mod_plat_ndh <- felm(log(Global_Sales) ~ Shooter |
                          Platform, vgsales_ndh)
summary(vg_mod_plat_ndh)

vg_mod_time_ndh <- felm(log(Global_Sales) ~ Shooter |
                          Year, vgsales_ndh)
summary(vg_mod_time_ndh)

vg_mod_reg_ndh <- felm(log(Global_Sales) ~ Shooter |
                         Region, vgsales_long_ndh)
summary(vg_mod_reg_ndh)

vg_mod_full_ndh <- felm(log(Global_Sales) ~ Shooter |
                          Publisher + Platform + Year + Region,
                        vgsales_long_ndh)
summary(vg_mod_full_ndh)
```

The `log(Global_Sales)` simple linear regression model has a `Shooter1` coefficient of about 0.334, which is similar to the models with fixed effects for Region, Year, and to a lesser extent Platform. However, including Publisher as an individual fixed effect or as part of the full fixed effects model produces a coefficient smaller by roughly half, indicating that the simple model is heavily biased upward. This demonstrates that including publisher fixed effects in the model is necessary to avoid omitted variable bias.

Further adjustments, such as the inclusion of a variable controlling for the number of games published per year, may introduce collinearity and thus were not tested for. This specific example was already controlled for by the Year fixed effect, which accounted for the number of games published per year as well as all other factors varying at the year level.

## Conclusion and Causality:

Regressions with Publisher, Platform, Year, and Region respectively held as fixed effects produce results very similar to the simple linear regression, suggesting that little omitted variable bias arises from the exclusion of these specific factors in the base model. A model containing all of these factors as fixed effects produces results less similar to the simple model and the individual fixed effects models, but still at a fairly nonsubstantial magnitude, reinforcing these conclusions. However, log-transformed models with the Duck Hunt outlier removed were heavily biased when not including Publisher as a fixed effect, indicating that of these variables, Publisher is a necessity to include in any model for analyzing Global Sales.

One important limitation that remains is the potential omission of game quality, an unobserved characteristic that is likely correlated with both genre and sales. Shooter games often share similar production values, design conventions, and franchise‑driven development cycles, which may make their average quality more homogeneous relative to other genres.If quality systematically differs between Shooter and non‑Shooter titles—and is not fully captured by publisher, platform, or year fixed effects—then the estimated Shooter coefficient may partially reflect these underlying quality differences rather than a pure genre effect. This remains a key source of unobserved heterogeneity that cannot be fully addressed with the available data.